Solution of hydrodynamic equilibrium in black-oil reservoirs: the classical material balance equation

### 2014-09-29 pituba/article/article_spe.tex
new changes

% Review
\documentclass[authoryear,preprint,review,11pt]{elsarticle}
%\documentclass[final,authoryear,5p,twocolumn,10pt]{elsarticle}

@@

\begin{abstract}
This work examines underlying details in obtaining the classical material balance equation from a three dimensional black-oil model typical of reservoir simulations. The three-dimensional model is integrated in reservoir volume and quantities such as fluid properties are spatially averaged. The material balance equation is obtained for averaged quantities and it is applied to calculate reservoir pressure. It is shown that the calculated pressure corresponds to the local pressure at an arbitrary reference depth and not to the average reservoir pressure in general. The definition of a convenient reference depth is discussed. It is found that the physical correctness of PVT properties obtained from laboratory experiments or empirical correlations is a sufficient condition for solution uniqueness of the material balance equation when pressure is the unknown quantity.
Finally, a numerical scheme for solving the material balance equation is proposed.
\end{abstract}

@@

It follows from Duhem's theorem that the equilibrium of oil and gas phases is determined if, in addition to the mass balance equations, two properties are prescribed. From Gibbs' phase rule, the system has two degrees of freedom. It is then necessary to specify two intensive properties. The constant temperature is one and pressure is chosen as the other. Oil phase density now becomes a function of pressure alone. The same is valid for mass fractions of gas and oil components in oil phase:
\begin{equation}
Y_g = Y_g (p_o) \, ,
\end{equation}
\begin{equation}
Y_o = Y_o (p_o) \, .
\end{equation}

@@

Eq~\eqref{eq: MBE} involves fluid properties, original fluid volumes ($N$, $W$ and $G)$, cumulative production and injection and aquifer influx.
It is an algebraic equation that can be solved for any one of its terms, provided that all remaining are known. This works focuses on its solution for pressure.

@@

The material balance equation is widely used to relate production and injection to reservoir pressure. Actually, all averaged fluid properties in Eq.~\eqref{eq: MBE} depend on the pressure field, and so does the pore volume. If the reservoir is under production, though, the pressure field is not known and for its prediction one would have to resort to more advanced techniques such as reservoir simulation.

@@

It can be shown that neglecting the variation of density with depth does not introduce any important error. As pressure increases with depth according to the hydrostatic gradient, density also increases but by a negligible amount. For instance, the average oil density can be approximated to the local density at the reference level: $\bar{\rho}_o (t) \approx \rho_o (t, z_{ref})$.

@@

Assuming reservoir temperature to be sufficiently above gas critical temperature for a negligible variation of the Z-factor with pressure, we use hydrostatic equilibrium again and obtain the derivative of gas density with depth:
\begin{equation}\label{eq: gradgdz}
\frac{1}{\rho_g}\left(\frac{\partial \rho_g}{\partial z}\right)_{T} = \frac{\rho_g \vec{g}}{p_g} = \left(\frac{\mathcal{M}}{ Z R T}\right) g \, .
\end{equation}

@@

For the case of an oil reservoir with a gas cap, the reference depth should be defined at the gas-oil contact. The point here is that at gas-oil contact both gas and oil densities are computed using the same pressure its local value. If the reference depth, common to all phases, is otherwise defined below gas-oil contact, gas density will increase with depth at a higher rate because of the oil hydrostatic gradient,
\begin{equation}
\frac{1}{\rho_g}\left(\frac{\partial \rho_g}{\partial z}\right)_{T,Z} = \frac{\rho_o}{\rho_g}\left(\frac{\mathcal{M}}{Z R T}\right)g  \, ,
\end{equation}
what is $\rho_o / \rho_g$ times greater than the gas hydrostatic gradient used in Eq.~\eqref{eq: gradgdz}.

It has been shown that the pressure variation with depth is not important for the computation of gas, oil and water densities. Nevertheless, pressure also plays a second role in the material balance equation by defining phase equilibrium, for what the pressure variation with depth may not be negligible in the solubility ratio: 
\begin{equation}\label{eq: dRsdz1}
\frac{1}{R_s}\frac{d R_s}{dz} = \frac{1}{R_s}\frac{d R_s}{dp_o} \frac{dp_o}{dz} = \frac{\rho_o g}{R_s}\frac{dR_s}{dp_o} \, .
\end{equation}

@@

In summary, the following four relations constitute sufficient conditions for $\varepsilon(p)$ be a strictly decreasing function and, therefore, for material balance equation to have only one solution:
\begin{equation}\label{eq: PVTH1}
\left(\frac{dB_o}{dp} - B_g \frac{dR_s}{dp} \right) \leq 0 \, ,
\end{equation}
\begin{equation}\label{eq: PVTH2}
\frac{dB_g}{dp} \leq 0 \, ,
\end{equation}
\begin{equation}\label{eq: PVTH3}
\frac{dB_w}{dp} \leq 0 \, ,
\end{equation}
\begin{equation}\label{eq: PVTH4}
\frac{dV_p}{dp} \geq 0 \, .
\end{equation}

Eq.~\eqref{eq: PVTH1} also provides a condition to check whether PVT data obtained from laboratory or empirical correlations is physically correct. It can be alternatively formulated in terms of densities as shown bellow in Eq.~\eqref{eq: vaporization}. It simply indicates that any vaporization must be followed by an increase in oil phase density:
\begin{equation}\label{eq: vaporization}
\frac{1}{\rho_o}\frac{d\rho_o}{dY_o} \geq Y_o \left( \frac{\rho_o}{\rho_g} - 1 \right) \, .
\end{equation}

@@

An open source implementation of the secant method is available in SciPy library \citep{scipy}. This work used library version $0.12$.

@@

\section{Results}
Material balance equation were solved for a black-oil reservoir initially producing in depletion drive mechanism and later subjected to water injection. Fig.~\ref{fig: matbal_p} shows in the solid line curve the severe depletion during initial phases of production. In the dashed curve, Fig.~\ref{fig: matbal_p} shows the bubble-point pressure, indicating how far the saturated reservoir is from undersaturation and allowing for planning of pressure maintenance projects.

Water injection starts when oil recovery factor $(RF)$ is about $RF=0.08$ and it is maintained until the reservoir reaches a novel undersaturated state. Initially, the high compressibility of gas prevents pressure from raising and free gas phase is reduced because of redissolution. Finally, the reservoir reaches the undersaturated state and a steep pressure increase is seen in Fig.~\ref{fig: matbal_p} for a recovery factor about $RF=0.15$.

@@

Figure~\ref{fig: epsilon} shows the shape of residual function ($\epsilon$) when pressure is varying and all remaining terms are held constant. Each different solid line corresponds to a different instant of time and therefore different volumes of cumulative production and injection. The function root is the solution of material balance equation, Eq.~\eqref{eq: MBE_short}, and the non-differentiable point is the bubble-point pressure. Also, it is seen that function $\epsilon$ monotonically decreases with pressure, as theoretically shown before.

@@

\section*{Highlights}
\begin{enumerate}[1.]
\item The existing link between black-oil and material balance equation is examined in detail.
\item It is shown that gravity and capillary forces must be neglected and saturations must be constant in space so one can integrate black-oil equations in reservoir volume and obtain material balance equation.
\item Material balance equation are shown to have a unique solution for a compressible system.
\item Material balance equation are formulated for a variable bubble-point and they are used to a reservoir under pressure maintenance.
\end{enumerate}

>>>>

% Review
%\documentclass[authoryear,preprint,review,11pt]{elsarticle}
\documentclass[final,authoryear,5p,twocolumn,10pt]{elsarticle}

@@

\begin{abstract}
This work examines underlying details in obtaining the classical material balance equation from a three dimensional black-oil model typical of reservoir simulations. The black-oil equations are integrated in reservoir volume and quantities such as fluid properties are spatially averaged. The material balance equation is obtained for averaged quantities and it is applied to calculate reservoir pressure. It is shown that the calculated pressure corresponds to the local pressure at an arbitrary reference depth and not to the average reservoir pressure in general. The definition of a convenient reference depth is also discussed. It is found that the physical correctness of PVT properties
% obtained from laboratory experiments or empirical correlations
is a sufficient condition for solution uniqueness of the material balance equation when pressure is the unknown quantity.
Finally, a numerical scheme for solving the material balance equation is proposed based on nonlinear root finding algorithms.
\end{abstract}

@@

It follows from Duhem's theorem that the equilibrium of oil and gas phases is determined if, in addition to the mass balance equations, two other properties are prescribed. From Gibbs' phase rule, the system has two degrees of freedom. It is then necessary to specify two intensive properties. The constant temperature is one and pressure is chosen as the other. Oil phase density now becomes a function of pressure alone. The same is valid for mass fractions of gas and oil components in oil phase:
\begin{equation}
Y_g = Y_g (p_o) \, ,
\end{equation}
\begin{equation}
Y_o = Y_o (p_o) \, .
\end{equation}

@@

Eq~\eqref{eq: MBE} involves fluid properties, original phase volumes ($N$, $W$ and $G)$, cumulative production and injection and aquifer influx.
It is an algebraic equation that can be solved for any one of its terms, provided that all remaining are known. This works focuses on its solution for pressure.

@@

The material balance equation is widely used to relate production and injection to reservoir pressure. Actually, all averaged phase properties in Eq.~\eqref{eq: MBE} depend on the pressure field, and so does the pore volume. If the reservoir is under production, though, the pressure field is not known and for its prediction one would have to resort to more advanced techniques such as reservoir simulation.

@@

It can be shown that neglecting the variation of density with depth does not introduce  important error. As pressure increases with depth according to the hydrostatic gradient, density also increases but by a negligible amount. For instance, the average oil density can be approximated to the local density at the reference level: $\bar{\rho}_o (t) \approx \rho_o (t, z_{ref})$.

@@

Assuming reservoir temperature to be sufficiently above gas critical temperature and thus a negligible variation of the Z-factor with pressure, we use hydrostatic equilibrium again and obtain the derivative of gas density with depth:
\begin{equation}\label{eq: gradgdz}
\frac{1}{\rho_g}\left(\frac{\partial \rho_g}{\partial z}\right)_{T} = \frac{\rho_g \vec{g}}{p_g} = \left(\frac{\mathcal{M}}{ Z R T}\right) g \, .
\end{equation}

@@

For the case of an oil reservoir with a gas cap, the reference depth should be defined at the gas-oil contact. The point here is that at gas-oil contact both gas and oil densities are computed using the pressure local value. If the reference depth, common to all phases, is otherwise defined below gas-oil contact, a higher gas density will be computed because of the oil hydrostatic gradient:
\begin{equation}
\frac{1}{\rho_g}\left(\frac{\partial \rho_g}{\partial z}\right)_{T,Z} = \frac{\rho_o}{\rho_g}\left(\frac{\mathcal{M}}{Z R T}\right)g  \, ,
\end{equation}
what is $\rho_o / \rho_g$ times greater than the gas hydrostatic gradient used in Eq.~\eqref{eq: gradgdz}.

It was shown that the pressure variation with depth is not important for the computation of gas, oil and water densities. Nevertheless, pressure also plays a second role in the material balance equation by defining phase equilibrium, for what the pressure variation with depth may not be negligible in the solubility ratio: 
\begin{equation}\label{eq: dRsdz1}
\frac{1}{R_s}\frac{d R_s}{dz} = \frac{1}{R_s}\frac{d R_s}{dp_o} \frac{dp_o}{dz} = \frac{\rho_o g}{R_s}\frac{dR_s}{dp_o} \, .
\end{equation}

@@

In summary, the following four relations constitute sufficient conditions for $\varepsilon(p)$ be a strictly decreasing function and, therefore, for material balance equation to have only one solution:
\begin{equation}\label{eq: PVTH1}
\left(\frac{dB_o}{dp} - B_g \frac{dR_s}{dp} \right) < 0 \, ,
\end{equation}
\begin{equation}\label{eq: PVTH2}
\frac{dB_g}{dp} < 0 \, ,
\end{equation}
\begin{equation}\label{eq: PVTH3}
\frac{dB_w}{dp} < 0 \, ,
\end{equation}
\begin{equation}\label{eq: PVTH4}
\frac{dV_p}{dp} > 0 \, .
\end{equation}

Eq.~\eqref{eq: PVTH1} also provides a condition to check whether PVT data obtained from laboratory or empirical correlations is physically correct. It can be alternatively formulated in terms of densities as shown bellow in Eq.~\eqref{eq: vaporization}. It simply indicates that any vaporization must be followed by an increase in oil phase density:
\begin{equation}\label{eq: vaporization}
\frac{1}{\rho_o}\frac{d\rho_o}{dY_o} > Y_o \left( \frac{\rho_o}{\rho_g} - 1 \right) \, .
\end{equation}

@@

An open source implementation of the secant method is available in SciPy library \citep{scipy}. This work used the library version $0.12$.

@@

\section{Results}
For illustrating the numerical procedure, the material balance equation was solved for a fictitious black-oil reservoir initially producing in a pure depletion drive mechanism and later subjected to water injection. Fig.~\ref{fig: matbal_p} shows in the solid line curve the severe depletion during initial phases of production. In the dashed curve, Fig.~\ref{fig: matbal_p} shows the bubble-point pressure, indicating how far the saturated reservoir is from undersaturation and allowing for planning of pressure maintenance projects.

Water injection starts when oil recovery factor $(RF)$ is $RF=0.08$ and it is maintained until the reservoir reaches a novel undersaturated state. Initially, the high compressibility of gas prevents pressure from raising and free gas phase is reduced because of redissolution. Finally, the reservoir reaches the undersaturated state and a steep pressure increase is seen in Fig.~\ref{fig: matbal_p} for a recovery factor about $RF=0.15$.

@@

Figure~\ref{fig: epsilon} shows the shape of residual function ($\epsilon$) when pressure is varying and all remaining terms are held constant. Each different solid line corresponds to a different instant of time and, therefore, to different volumes of cumulative production and injection. The function root is the solution of material balance equation, Eq.~\eqref{eq: MBE_short}, and the non-differentiable point is the bubble-point pressure. Also, it is seen that function $\epsilon$ monotonically decreases with pressure, as theoretically shown before.

@@

%\section*{Highlights}
%\begin{enumerate}[1.]
%\item The existing link between black-oil and material balance equation is examined in detail.
%\item It is shown that gravity and capillary forces must be neglected and saturations must be constant in space so one can integrate black-oil equations in reservoir volume and obtain material balance equation.
%\item Material balance equation are shown to have a unique solution for a compressible system.
%\item Material balance equation are formulated for a variable bubble-point and they are used to a reservoir under pressure maintenance.
%\end{enumerate}
